A Java Framework For Embarrassingly Parallel Problems

### 2013-11-29 paper/paper.tex
More paper work

\doublespacing
\begin{document}
\title{A Java Wrapper For Embarrassingly Parallel Problems}
\author{Jacob Schwartz}
\maketitle

@@

% TODO: Arrows?
The motivation for this effort was based on research into different forms of the
species \emph{Cicer arietinum}, the chickpea. The two main forms, Kabuli and
Desi, are very different in size and color. The Desi is also more resistant to
diseases and are more effective at root nodule symbiosis. The Kabuli has a large
but soft shell, which make it at better food source for humans. Both forms are
cultivated and have been under human selection for several thousand years. We
wanted to use an existing Ka/Ks program (Zhang et al. \cite{kaks}) to quantify
the evolution and selection of these chickpeas. We needed to analyze three
variants, each with thousands of sequences. The Ka/Ks Calculator program took
between five to thirty seconds to process one pair of sequences on a ordinary
machine.

@@

%TODO: get a better source? Talk about UCLUST and octupus
% Talk about uclust and octupus and blast, even though there are parallelize
% versions. Rewrite to not go into details about the programs but to talk about
% that there are many programs.
Ka/Ks Calculator is not the only embarrassingly parallel program that has the
potential to take several days to run on a data set. In fact, there are several
embarrassingly parallel programs in the biological sciences. To deal with large
amounts of data, the best method is often to group them into clusters.
Hierarchical clustering algorithms take the data and then organize into
hierarchical structure, usually a dendrogram, based on a priority matrix. There
are hierarchical clustering algorithms that can be implemented in parallel using
an embarrassingly parallel approach.  With several applications that would
benefit from parallelism in this field, it was obvious it would be desirable to
develop a framework for the parallelization task. \cite{cluster}

%TODO: Footnotes for days
This framework is designed to run locally but there are other options when more
power is needed. Gunarathne et al.\cite{cloud} have investigated running much
larger embarrassingly parallel biological application using various cloud
services. These services offer on-demand computational power with flexibility.
These services have built in hooks to various distributed services, such as
storage and databases. Amazon has a set of cloud computing services called
Amazon Web Services(AWS) and Microsoft has a similar service named Microsoft
Azure. AWS and Azure are esentially virtual instances run on machines in the
cloud that can be spun up for extra computing power and shut down when not
needed.

Amazon also offers another service called Amazon Elastic MapReduce. This
combines its virtual instance framework with Apache Hadoop. As seen
in figure \ref{fig:map}, Hadoop breaks down the input data into parts, runs them in parallel
with the 'Map' function and then combines the results with the 'Reduce'
function. Dryad is a framework developed released by Microsoft Research that
allows the end-user to create directed acyclic computation graphs that run on
Windows High Performance Computer(HPC) clusters. Gunarathne et al. compared
the services using several applications to analyze performance, cost and
usability.

@@

We evaluated the convenience and performace of the framework on two different
applications: the Ka/Ks Calculator and two different sequence clustering
algorithms. In all cases, we compared the framework based application with the
original using different data partitions and different thread counts.

\subsection{Ka/Ks Calculator}
Ka/Ks Calculator was run through the framework with ten different samples, each 
containing sixteen units of work. Each sample was run by the framework thread
counts, starting at 1 and increasing by 
powers of two until the 100\% processor usage (16 processors). These tests were run
with minimal other activity on the server.

@@

In the future, we would like to see this work be split into a
local front-end and a cloud back-end. This allows for an even greater runtime
decrease because the number of compute nodes available is larger than the number
of processors on most machines. Splitting the framework also allows for better
testing of cloud environments, where the cloud environment itself could be a
configurable option for end users.
TODO: MORE

>>>>

\doublespacing
\begin{document}
\title{A Java Framework For Embarrassingly Parallel Problems}
\author{Jacob Schwartz}
\maketitle

@@

The motivation for this effort was based on research into different forms of the
species \emph{Cicer arietinum}, the chickpea. The two main forms, Kabuli and
Desi, are very different in size and color. The Desi is also more resistant to
diseases and is more effective at root nodule symbiosis. The Kabuli has a large
but soft shell, which makes it a better food source for humans. Both forms are
cultivated and have been under human selection for several thousand years. We
wanted to use an existing Ka/Ks program (Zhang et al. \cite{kaks}) to quantify
the evolution and selection of these chickpeas. We needed to analyze three
variants, each with thousands of sequences. The Ka/Ks Calculator program took
between five to thirty seconds to process one pair of sequences on a ordinary
machine.

@@

Ka/Ks Calculator is not the only embarrassingly parallel program that has the
potential to take several days to run on a data set. In fact, there are several
embarrassingly parallel programs in the biological sciences. Basic Local
Alignment Search Tool, or BLAST\fotenote{blast.ncbi.nlm.nih.gov}, is an
algorithm that compares many aspects of DNA sequences, like the nucleotides or
amino acid sequences. In addition, the user can also compare to the sequences to
a database of sequences to identity known subsequences. There are many
implementations of BLAST, some of them already have parallel implementations.
Many applications however do not have parallel processing capabilities. A class
of problems that organize large amounts of data into similar groups called
clusters. Two applications, octupus\footnote{octupus.sourceforge.com} and
USEARCH\footnote{https://www.drive5.com/usearch}, are examples of programs in
this class that do not have parallel versions. With several applications that
would benefit from parallelism in this field, it was obvious it would be
desirable to develop a framework for the parallelization task.

This framework is designed to run locally but there are other options when more
power is needed. Gunarathne et al.\cite{cloud} have investigated running much
larger embarrassingly parallel biological application using various cloud
services. These services offer on-demand computational power with flexibility.
These services have built in hooks to various distributed services, such as
storage and databases. Amazon has a set of cloud computing services called
Amazon Web Services(AWS)\footnote{https://aws.amazon.com} and Microsoft has a
similar service named Microsoft Azure\footnote{https://www.windowsazure.com}.
AWS and Azure are esentially virtual instances run on machines in the cloud that
can be spun up for extra computing power and shut down when not needed.

Amazon also offers another service called Amazon Elastic
MapReduce\footnote{https://aws.amazon.com/elasticmapreduce}. This combines its
virtual instance framework with Apache
Hadoop\footnote{https://hadoop.apache.org}. As seen in figure \ref{fig:map},
Hadoop breaks down the input data into parts, runs them in parallel with the
'Map' function and then combines the results with the 'Reduce' function.
DryadLINQ\footnote{https://research.microsoft.com/apps/mobile/showpage.aspx?page=/en-us/projects/draydlinq}
is a framework developed released by Microsoft Research that allows the end-user
to create directed acyclic computation graphs that run on Windows High
Performance Computer(HPC)\footnote{https://www.microsoft.com/hpc} clusters.
Gunarathne et al. compared the services using several applications to analyze
performance, cost and usability.

@@

We evaluated the convenience and performace of the framework on two different
applications: the Ka/Ks Calculator and two different sequence clustering
algorithms, USEARCH and ocutupus. In all cases, we compared the framework based
application with the original using different data partitions and different
thread counts.

\subsection{Ka/Ks Calculator}
TODO: Reword when redone
Ka/Ks Calculator was run through the framework with ten different samples, each
containing sixteen units of work. Each sample was run by the framework thread
counts, starting at 1 and increasing by powers of two until the 100\% processor
usage (16 processors). These tests were run with minimal other activity on the
server.

@@

In the future, we would like to see this work be split into a local front-end
and a cloud back-end. The configuration set-up process will run locally. The
split and merge steps depend on the cloud service chosen, but data processing
will always occur in the cloud. Splitting the framework will give performance
improvements, because the number of compute nodes can be much larger than the
number of processors on an average machine. More customization options would be
available to the end-user, such as choice of environment. This provides ultimate
control for the group to pick the best enviroment to suit its computational and
financial needs.
